---
title: "Data Cleaning"
author: "Chris Okitondo & Irene Cavros"
date: "2023-02-20"
output: html_document
---


# Setup

Load needed packages. make sure they are installed.

```{r, warning=FALSE, message=FALSE}
library(readxl) #for loading Excel files
library(writexl) #for writing Excel files
library(haven) # for loading SPSS file
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(tidyverse)
```


# Data loading
```{r}
data_location <- here::here("data","raw_data","nhis_00015.sav")
rawdata <- haven:: read_sav(data_location)
```


# Check data

We are checking the data using several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```


### Observations for each year: 2012-2021
```{r}
table(rawdata$YEAR)
```





# Variables for Analysis

Demographics: age, sex, race, ethnicity, veteran status
Socio-economic status: education level, total combined family income, employment status, paid sick leave as part of job benefits, ratio of family income to poverty threshold
General health: perceived health status, categorical BMI
Conditions: ever told had asthma, ever told had cancer, ever told had coronary heart disease
Health behaviors: alcohol drinking status, cigarette smoking status
Access to care: has usual place for medical care, medical care delayed due to cost (past 12 months)
Health insurance: health insurance coverage status
Mental health: anxiety level, depression level
Flu vaccine: Had any flu vaccine, past 12 months.



# Cleaning varibles to create appropriate levels and also exclusing missing or unknown values
```{r}
# Age
rawdata$AGE[rawdata$AGE == 997 | rawdata$AGE == 998 | rawdata$AGE == 999] <- NA

# Sex: 1 for Male and 2 for female, 7 or 9 for unknown
rawdata$SEX[rawdata$SEX == 7 |  rawdata$SEX == 9] <- NA


# White:1, Black:2, American Indiana/Alaska Native:3, Asian: 4
rawdata$RACENEW[rawdata$RACENEW == 100] <- 1
rawdata$RACENEW[rawdata$RACENEW == 200] <- 2
rawdata$RACENEW[rawdata$RACENEW == 300] <- 3
rawdata$RACENEW[rawdata$RACENEW == 400] <- 4
rawdata$RACENEW[rawdata$RACENEW > 400] <- NA

# Hispanic: 1 for no, 2 for yes.
rawdata$HISPYN[rawdata$HISPYN == 7 | rawdata$HISPYN == 8 | rawdata$HISPYN == 9] <- NA

# Veteran status: 1 for no. 2 for yes
rawdata$ARMFEV[rawdata$ARMFEV==10|rawdata$ARMFEV == 11 |rawdata$ARMFEV == 12 ] <- 1
rawdata$ARMFEV[rawdata$ARMFEV == 20]  <- 2
rawdata$ARMFEV[rawdata$ARMFEV==0|rawdata$ARMFEV==97|rawdata$ARMFEV ==98|rawdata$ARMFEV == 99 ]<-NA


# EDUCATION: 1 for less than HS, 2 for HS, 3 for Some college, 4 for bachelor, 5 for graduate degree
rawdata$EDUC[rawdata$EDUC >= 102 & rawdata$EDUC <= 116] <- 1
rawdata$EDUC[rawdata$EDUC >= 200 & rawdata$EDUC <= 202] <- 2
rawdata$EDUC[rawdata$EDUC >= 300 & rawdata$EDUC <= 303] <- 3
rawdata$EDUC[rawdata$EDUC == 400] <- 4
rawdata$EDUC[rawdata$EDUC >= 500 & rawdata$EDUC <= 505] <- 5
rawdata$EDUC[rawdata$EDUC > 505] <- NA


# EMPLOYMENT STATUS: 1 for not employed, 2 for employed, and 3 for retired
rawdata$EMPSTAT[rawdata$EMPSTAT >= 100 &  rawdata$EMPSTAT <=120] <- 1
rawdata$EMPSTAT[rawdata$EMPSTAT == 200] <- 2
rawdata$EMPSTAT[rawdata$EMPSTAT == 220] <- 3

# PAID Sick: 1 for no, 2 for yes


# INCOME: 0 - 49,999: 1, 50,000-99,999: 2, AND +100k: 3
rawdata$INCFAM07ON[rawdata$INCFAM07ON >= 10 &  rawdata$INCFAM07ON <= 12] <- 1
rawdata$INCFAM07ON[rawdata$INCFAM07ON >= 22 &  rawdata$INCFAM07ON <= 23] <- 2
rawdata$INCFAM07ON[rawdata$INCFAM07ON == 24] <- 3
rawdata$INCFAM07ON[rawdata$INCFAM07ON > 24] <- NA
rawdata$INCFAM07ON[rawdata$INCFAM07ON == 20 | rawdata$INCFAM07ON == 21] <- NA


# HEALTH STATUS: 1 Excellent, 2: Very good, 3: Good, 4: Fair, 5: Poor
rawdata$HEALTH [rawdata$HEALTH  > 5] <- NA

# BMI CAT: 1 for BMI < 18.5 (Underweight), 2 for 18.5 - 25 (Normalt), 3 for 25 - 30 (Overweight), 4 for >30 (Obese)
rawdata$BMICAT [rawdata$BMICAT  == 0 | rawdata$BMICAT  == 9] <- NA

# USUAL PLACE FOR MEDICAL CARE: 1 for no, 2 and 3 for yes
rawdata$USUALPL[rawdata$USUALPL == 1] <- 1
rawdata$USUALPL[rawdata$USUALPL == 2 |  rawdata$USUALPL == 3] <- 2
rawdata$USUALPL[rawdata$USUALPL==0 | rawdata$USUALPL==7| rawdata$USUALPL==8| rawdata$USUALPL==9] <- NA

# Medical care delayed due to cost: 1 for no, 2 for yes
rawdata$USUALPL[rawdata$DELAYCOST ==0 | rawdata$DELAYCOST ==7| rawdata$DELAYCOST ==8| rawdata$USUALPL==9] <- NA

# Health insurance coverage: 1 for no, 2 for yes
rawdata$HINOTCOVE[rawdata$HINOTCOVE == 2] <- 3
rawdata$HINOTCOVE[rawdata$HINOTCOVE == 1] <- 2
rawdata$HINOTCOVE[rawdata$HINOTCOVE == 3] <- 1
rawdata$HINOTCOVE[rawdata$HINOTCOVE == 9] <- NA

# Had flu vaccine, past 12 months
rawdata$VACFLU12M [rawdata$VACFLU12M  == 0|rawdata$VACFLU12M  == 3 |rawdata$VACFLU12M ==7|rawdata$VACFLU12M ==8|rawdata$VACFLU12M ==9] <- NA

# Had covid, according to test
rawdata$CVDTESTRSLT [rawdata$CVDTESTRSLT  == 0|rawdata$CVDTESTRSLT  == 3 |rawdata$CVDTESTRSLT ==7|rawdata$CVDTESTRSLT ==8|rawdata$CVDTESTRSLT ==9] <- NA

# Ever told has Asthma: ASTHMAEV: 1 for no, 2 for yes
rawdata$ASTHMAEV[rawdata$ASTHMAEV == 0|rawdata$ASTHMAEV==7|rawdata$ASTHMAEV==9] <- NA

# Ever told has Cancer: CANCEREV: 1 for no, 2 for yes
rawdata$CANCEREV[rawdata$CANCEREV == 0|rawdata$CANCEREV==7|rawdata$CANCEREV==9] <- NA


# Ever told has CHD: CHEARTDIEV 1 for no, 2 for yes
rawdata$CHEARTDIEV[rawdata$CHEARTDIEV==0|rawdata$CHEARTDIEV==7|rawdata$CHEARTDIEV==9] <- NA


# ALCSTAT1: Alcohol drinking: 1 for no drinker, 2 for former drinker, 3 for current drinker
rawdata$ALCSTAT1 [rawdata$ALCSTAT1 ==0|rawdata$ALCSTAT1 ==9] <- NA


# Smoking status: 1 for no smoker, 2 for former smoker, 3 for current smoker
rawdata$SMOKESTATUS2[rawdata$SMOKESTATUS2 == 10|rawdata$SMOKESTATUS2  == 11|rawdata$SMOKESTATUS2==12] <-3
rawdata$SMOKESTATUS2 [rawdata$SMOKESTATUS2 == 20 ] <- 2
rawdata$SMOKESTATUS2 [rawdata$SMOKESTATUS2 == 30 ] <- 1
rawdata$SMOKESTATUS2 [rawdata$SMOKESTATUS2 > 30 ] <- NA

# Worry, nervous, or anxious frequency: WORFREQ. 1 Daily, 2 weekly, 3 Monthly, 4 a few times a year, 5 Never
rawdata$WORFREQ [rawdata$WORFREQ ==0|rawdata$WORFREQ ==7|rawdata$WORFREQ ==8
                |rawdata$WORFREQ ==9] <- NA

# Depression frequency: DEPFREQ. 1 Daily, 2 weekly, 3 Monthly, 4 a few times a year, 5 Never
rawdata$DEPFREQ[rawdata$DEPFREQ==0|rawdata$DEPFREQ==7|rawdata$DEPFREQ==8|rawdata$DEPFREQ==9] <- NA
```



# Viewing data
```{r}
View(rawdata)
```


# Removing NAs values: We will decide how to proceed.
```{r}
#rawdata <- na.omit(rawdata)
```


# Rename some variables
```{r}
rawdata <- rename(rawdata, age = AGE, sex = SEX, race = RACENEW, ethnicity = HISPYN, veteran_status = ARMFEV, education = EDUC, employment_status = EMPSTAT, paidsick = PAIDSICK, fam_income = INCFAM07ON, health_status = HEALTH, bmi = BMICAT, medical_care_access = USUALPL, medical_care_delayed_cost = DELAYCOST, health_ins_status = HINOTCOVE, asthma = ASTHMAEV, cancer = CANCEREV, chd = CHEARTDIEV, alcohol_status = ALCSTAT1, smoking_status = SMOKESTATUS2, worriedness	= WORFREQ, depression = DEPFREQ, flu_vaccine = VACFLU12M, covid = CVDTESTRSLT)
```


### Check new variable names
```{r}
names(rawdata)
```

### Converting some variables to characters
```{r}
rawdata <- rawdata %>% 
     mutate_at(c(8:30), as.character)
```

### Creating a copy of rawdata
```{r}
processeddata <- rawdata
```

### saving the processed data
```{r}
save_data_location <- here::here("data","processed_data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```


